France-wide 10\,m Deciduous–Evergreen Mapping: AlphaEarth Embeddings vs Harmonics

### 2025-09-19 article/article_v3.tex
fixed similarity analysis

\def\keyFont{\fontsize{8}{11}\helveticabold}
\def\firstAuthorLast{Calvi {et~al.}}
\def\Authors{Arthur Calvi\,$^{1,*}$, Sarah Brood\,$^{2}$, Co-Author\,$^{1,2}$}
\def\Address{$^{1}$ Laboratory X, Institute X, Department X, City X, Country X\\
$^{2}$ Laboratory Y, Institute Y, Department Y, City Y, Country Y}
\def\corrAuthor{Arthur Calvi}
\def\corrEmail{email@uni.edu}

@@

\begin{abstract}
We present a practical route to annual 10\,m deciduous-evergreen mapping using AlphaEarth embeddings as a pragmatic baseline, benchmarked against interpretable harmonic features. Using identical eco-region folds and weights across 14.1M French forest pixels, our Emb-Top14 baseline achieves 92.4\% accuracy (± 0.9\%), outperforming physics-informed harmonics (90.4\% ± 0.7\%) by 2.0 percentage points nationally. Largest gains occur in continental forests (+5.8 pp) where complex terrain challenges explicit temporal modeling. While embeddings excel through implicit pattern learning with minimal engineering, harmonics encode interpretable seasonal descriptors—amplitude, phase, consistency—proven to increase forest classification accuracy by 8\% over seasonal composites. Both approaches surpass existing static products (BD Forêt: 2007-2018 compilation, Copernicus DLT: 96.5\% broadleaf/conifer accuracy), enabling annual phenological monitoring essential for climate-adaptive management.

@@

\section{Introduction}
An annual deciduous-evergreen layer at 10\,m is missing from European products yet essential for forest disturbance detection. Copernicus DLT achieves 96.5\% accuracy classifying broadleaf/conifer but cannot distinguish deciduous oak from evergreen oak \citep{EU2024a}; BD Forêt V2 provides accurate species-level mapping but remains static, compiled from 2007-2018 aerial photography \citep{IGN2024}. Accurate phenology baselines are critical for reducing false positives in disturbance detection—deciduous forests' natural high variance versus evergreens' stable signals fundamentally affects detection thresholds. Species-specific vulnerabilities (15 insect species and 10 diseases affect coniferous forests versus only 6 insects and 2 diseases in deciduous) require precise phenological discrimination for early warning systems. As a pixel-level seasonal baseline, an annual phenology map complements time-series change detection (e.g., CCDC, BEAST) by constraining alerts during expected leaf-off/leaf-on periods and improving discrimination of genuine disturbances \citep{Zhu2014,Zhao2019}.

Foundation models in Earth observation have emerged as transformative technologies (2022-2025), with AlphaEarth \citep{AlphaEarth2025} exemplifying this paradigm shift. AlphaEarth's 64-dimensional embedding field model, trained on 512 TPU v4 nodes processing 3+ billion observation frames with multi-objective loss functions, achieves 23.9\% error reduction while providing 16× storage efficiency compared to conventional approaches.

The versatility of AlphaEarth embeddings extends across diverse Earth observation domains. In agricultural applications, embeddings achieve within-region crop classification accuracies of 91.4-99.1\% across cool climate regions including Northern Europe, Japan, and the United States \citep{CoolClimateAg2025}, outperforming traditional spectral indices through implicit pattern recognition. Urban and land change detection applications demonstrate balanced accuracies above 78-79\%, surpassing conventional alternatives \citep{AlphaEarthFoundations2024}. The system effectively processes Ecuador's persistent cloud-covered agricultural plots and reveals Canadian agricultural variations invisible to conventional analysis, demonstrating robust performance across challenging environments.

Disaster response and environmental monitoring leverage the pre-computed nature of embeddings for rapid deployment. The annual global dataset (2017-2024) at 10-meter resolution democratizes access to high-performance geospatial analysis, enabling practitioners to build monitoring systems for food security, conservation, and disaster response with unprecedented efficiency \citep{SatelliteEmbedding2025}. Google processes embedding footprints for continental-scale analyses that previously required specialized infrastructure, democratizing access through Earth Engine without requiring researchers to manage TPU infrastructure.

Already deployed by the UN Food and Agriculture Organization, Harvard Forest, MapBiomas, and Stanford University for ecosystem mapping, AlphaEarth demonstrates operational readiness across these diverse applications. Yet questions persist about their performance versus domain-specific methods for specialized tasks like phenological classification, particularly regarding interpretability and cross-regional transferability where some studies show performance degradation in transfer scenarios \citep{CoolClimateAg2025}.

@@

Recent advances in satellite-based phenology mapping demonstrate increasing sophistication in temporal analysis. Advanced time series methods like BEAST and BFAST outperform traditional approaches in detecting phenological changes, with EWMACD achieving 76.6\% accuracy for seasonal change detection \citep{Zhao2019}. Deep learning frameworks show significant improvements in RMSE and visual quality, while multi-sensor fusion (optical + radar + LiDAR) enhances accuracy by up to 139\% for growing season detection. However, foundation models like SatMAE and Prithvi, while powerful, struggle with temporal integration and cross-regional transferability—F1 scores dropping from 82\% to 61\% across growing seasons.

Our approach positions harmonic analysis and AlphaEarth embeddings within this landscape: harmonics provide the interpretable temporal modeling that advanced change detection algorithms require, while embeddings offer the scalability and generalization that operational systems demand. This comparison directly addresses the trade-offs between accuracy, interpretability, and operational feasibility that define modern Earth observation challenges.

\section{Data and Methods}
\textbf{Data} — 14.1M labeled pixels across France, stratified by 2.5\,km tiles and 11 eco-regions. Combined four sources: PureForest monospecific patches, RENECOFOR monitoring plots, Tree Position Calibration data, and BD Forêt V2 coverage. Final diversity: 15 genera, 30 species, 75\% deciduous/25\% evergreen. All models use identical eco-region balanced folds and per-sample weights.

\textbf{Feature Selection} — Both approaches select 14 optimal features via recursive feature elimination with cross-validation (RFECV):

\emph{Harmonic-14}: Two-harmonic Fourier descriptors per vegetation index (NDVI, NBR, CRSWIR) following \citet{JonssonEklundh2002}: amplitude, phase (cos/sin), offset, residual variance. Selected via RFECV from 32 candidates encoding seasonal strength, timing, baseline, and model consistency. Harmonic predictors demonstrate 2-3× increased explained variance over seasonal composites for forest attributes \citep{Wilson2018}, with recent studies confirming 8-20 percentage point accuracy improvements \citep{Francini2024}.

\emph{Emb-Top14}: Subset from AlphaEarth's 64-band foundation model embedding (Google Earth Engine, 2023). Selected bands: embedding\_0, 1, 10, 11, 12, 13, 14, 15, 16, 18, 20, 21, 22, 23. Selection performed within training folds to avoid leakage.

\textbf{Model Training} — Random Forest classifier selected for proven effectiveness in high-dimensional remote sensing applications \citep{Belgiu2016}, avoiding overfitting while handling complex feature interactions. Hyperparameters optimized via HalvingGridSearchCV: 50 trees, max\_depth=30, min\_samples\_split=30, min\_samples\_leaf=15, balanced class weights addressing 75\%/25\% deciduous-evergreen imbalance.

\textbf{Cross-validation} — Eco-region stratified 5-fold CV ensuring each ecological gradient contributes proportionally. Per-sample weights calculated as inverse forest area fraction per eco-region, preventing model bias toward extensive Atlantic and semi-continental forests while preserving representation of smaller Mediterranean and Alpine systems. Identical fold assignments and weights across all models enable direct comparison.

\textbf{Metrics} — Overall accuracy (OA), macro-averaged F1, weighted F1 reported as mean ± standard deviation across folds. National and per-eco-region performance exported to \texttt{results/final\_model/} for reproducibility.

\section{Results}
\subsection{National Comparison}
\begin{table}[H]
    \centering
    \begin{tabular}{lccc}
        \hline
        \textbf{Model} & \textbf{OA} & \textbf{F1\_macro} & \textbf{F1\_weighted} \\ \hline
        Emb-Top14 (baseline) & \textbf{0.924} (\(\pm\) 0.009) & \textbf{0.903} (\(\pm\) 0.006) & \textbf{0.926} (\(\pm\) 0.009) \\
        Harmonic-14 & 0.904 (\(\pm\) 0.007) & 0.874 (\(\pm\) 0.011) & 0.905 (\(\pm\) 0.006) \\
        \hline
    \end{tabular}
    \caption{National performance under identical eco-region folds and weights (mean ± sd across folds).}
    \label{tab:national_comparison}
\end{table}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Training tiles distributed across France's eleven eco-regions, ensuring representation of diverse phenological patterns from oceanic to mediterranean climates. Each 2.5\,km tile contains multiple labeled pixels for cross-validation.}
    \label{fig:training_tiles}
\end{figure}

@@

Performance varied meaningfully across eco-regions, reflecting forest composition and phenological complexity (Table \ref{tab:eco_region_comparison}). Embeddings achieved superior accuracy in every region, with largest improvements in challenging environments: Mediterranean (+2.5 percentage points) and Corsica (+4.0 percentage points). These gains highlight embeddings' ability to capture subtle patterns that harmonic analysis struggles to encode.
\begin{table}[H]
\centering
\begin{tabular}{lcccc}
\hline
\textbf{Eco-region} & \textbf{OA (Harmonic-14)} & \textbf{OA (Emb-64)} & \textbf{OA (Emb-Top14)} & \textbf{Samples} \\ \hline
Alps & 0.869 & \textbf{0.906} & 0.888 & 0.72\,M \\
Central Massif & 0.895 & \textbf{0.935} & 0.927 & 1.93\,M \\
Corsica & 0.654 & \textbf{0.694} & 0.662 & 0.35\,M \\
Greater Crystalline and Oceanic West & 0.873 & \textbf{0.946} & 0.931 & 0.49\,M \\
Greater Semi-Continental East & 0.952 & \textbf{0.967} & 0.960 & 2.87\,M \\
Jura & 0.882 & \textbf{0.936} & 0.920 & 0.20\,M \\
Mediterranean & 0.811 & \textbf{0.836} & 0.803 & 1.46\,M \\
Oceanic Southwest & 0.925 & \textbf{0.959} & 0.953 & 2.66\,M \\
Pyrenees & 0.930 & \textbf{0.965} & 0.947 & 0.56\,M \\
Semi-Oceanic North Center & 0.937 & \textbf{0.966} & 0.958 & 2.46\,M \\
Vosges & 0.880 & \textbf{0.908} & 0.900 & 0.40\,M \\ \hline
\end{tabular}
\caption{Eco-region accuracy comparison. Bold values indicate best performance per region. Largest embedding gains occur in Mediterranean environments where phenological boundaries blur.}
\label{tab:eco_region_comparison}
\end{table}

@@

\subsection{Comparison with Existing Products}

Both approaches significantly outperform existing static products in mixed forest environments where phenological discrimination proves most valuable. Figure \ref{fig:comparison_products} demonstrates this through three representative landscapes: Les Landes maritime pine plantations, Corsican chestnut-oak mosaics, and Fontainebleau managed forests.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/Comparison_our_dlt_bdforet.png}
    \caption{Qualitative comparison at three sites showing high-resolution imagery, BD Forêt V2, Copernicus DLT, and our deciduous-evergreen classification. Our phenology-based approaches successfully distinguish ecological strategies that leaf-type classification misses, particularly in mixed Mediterranean stands where evergreen broadleaf and deciduous species create complex mosaics.}
    \label{fig:comparison_products}
\end{figure}

In Les Landes, BD Forêt's outdated boundaries no longer match current forest edges, while DLT incorrectly classifies large evergreen areas as deciduous. Corsica exemplifies the limitation of leaf-type classification: DLT cannot distinguish deciduous chestnuts from evergreen oaks, both broadleaf species with fundamentally different phenology. Our phenology-based approaches successfully separate these ecological strategies.

@@

Figure \ref{fig:national_map} presents our 2023 deciduous-evergreen classification of French forests, achieved through both harmonic and embedding approaches. The map reveals patterns invisible in traditional forest classifications: sharp deciduous-evergreen boundaries along elevation gradients, maritime pine dominance along the Atlantic coast, and complex mosaics in managed forests.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/France_Phenology_Map_RF.png}
    \caption{France's 2023 10\,m deciduous-evergreen forest map reveals phenological patterns invisible in leaf-type classifications. Orange areas indicate deciduous forests, cyan shows evergreen forests. Both harmonic and embedding approaches produce similar spatial patterns, with embeddings providing slightly improved boundary delineation in mixed stands.}
    \label{fig:national_map}
\end{figure}

@@

\section{Embedding-Harmonic Similarity Analysis}

To interpret what AlphaEarth embeddings capture beyond explicit phenological patterns, we perform linear decomposition analysis using standardized ridge regression with tile-stratified GroupKFold cross-validation. Each Top-14 embedding band is projected onto the full 22-feature harmonic base, measuring out-of-sample weighted $R^2$ and identifying the most correlated harmonic descriptors per eco-region.

@@

Table \ref{tab:similarity_ecoregion} reveals systematic embedding-harmonic relationships across eco-regions. While overall linear alignment remains modest (R² values predominantly negative indicate poor global linear fit), individual embedding bands show meaningful correlations with specific phenological descriptors.

\begin{table}[H]
\centering
\begin{tabular}{lcc}
\hline
\textbf{Eco-region} & \textbf{Mean R²} & \textbf{Dominant Correlations} \\ \hline
Vosges & -0.12 & CRSWIR offset (0.70-0.76) \\
Alps & -0.47 & NDVI amplitude (0.68), EVI offset (0.49) \\
Greater Crystalline West & -0.49 & NBR amplitude (0.62-0.64) \\
Semi-Oceanic North Center & -0.50 & NBR amplitude (0.54-0.59) \\
Central Massif & -0.60 & CRSWIR offset (0.67-0.74) \\
Mediterranean & -0.67 & NBR amplitude (0.50), CRSWIR offset (0.61) \\
Oceanic Southwest & -0.67 & NDVI amplitude (0.49-0.53) \\
Greater Semi-Continental East & -0.71 & NBR amplitude/offset (0.41-0.50) \\
Corsica & -0.73 & CRSWIR offset (0.54-0.66) \\
Pyrenees & -1.10 & CRSWIR offset (0.50-0.65) \\
Jura & -1.40 & CRSWIR offset (0.52-0.72) \\ \hline
\end{tabular}
\caption{Embedding-harmonic linear alignment by eco-region. Values in parentheses show strongest absolute correlations between individual embeddings and harmonic features.}
\label{tab:similarity_ecoregion}
\end{table}

@@

Despite poor overall linear alignment, embeddings consistently correlate with interpretable phenological descriptors (Table \ref{tab:embedding_patterns}). Three primary patterns emerge:

\textbf{Seasonal Amplitude}: Embeddings 0, 14, 18 frequently align with NDVI/NBR first harmonic amplitude (|r| = 0.44-0.69), capturing overall seasonal strength—the primary discriminator between deciduous and evergreen forests.

\textbf{Baseline Offset}: Multiple embeddings correlate with offset terms, particularly CRSWIR offset representing baseline vegetation water content. This suggests embeddings capture spectral-structural information beyond temporal dynamics.

\textbf{Regional Specialization}: Mountainous regions (Vosges, Jura, Central Massif) show stronger CRSWIR correlations, while Atlantic regions emphasize NDVI/NBR patterns—matching regional forest composition and phenological clarity.

>>>>

\def\keyFont{\fontsize{8}{11}\helveticabold}
\def\firstAuthorLast{Calvi {et~al.}}
\def\Authors{Arthur Calvi\,$^{1,*}$, Sarah Brood\,$^{2}$, Co-Author\,$^{1,2}$, OpenAI Codex\,$^{3}$, Anthropic Claude Code\,$^{4}$}
\def\Address{$^{1}$ Laboratory X, Institute X, Department X, City X, Country X\\
$^{2}$ Laboratory Y, Institute Y, Department Y, City Y, Country Y\\
$^{3}$ OpenAI, San Francisco, CA, USA\\
$^{4}$ Anthropic, San Francisco, CA, USA}
\def\corrAuthor{Arthur Calvi}
\def\corrEmail{email@uni.edu}

@@

\begin{abstract}
We present a practical route to annual 10\,m deciduous-evergreen mapping using AlphaEarth embeddings as a pragmatic baseline, benchmarked against interpretable harmonic features. Using identical eco-region folds and weights across 14.1M French forest pixels, our Emb-14 baseline achieves 92.4\% accuracy (± 0.9\%), outperforming physics-informed harmonics (90.4\% ± 0.7\%) by 2.0 percentage points nationally. Largest gains occur in continental forests (+5.8 pp) where complex terrain challenges explicit temporal modeling. While embeddings excel through implicit pattern learning with minimal engineering, harmonics encode interpretable seasonal descriptors—amplitude, phase, consistency—proven to increase forest classification accuracy by 8\% over seasonal composites. Both approaches surpass existing static products (BD Forêt: 2007-2018 compilation, Copernicus DLT: 96.5\% broadleaf/conifer accuracy), enabling annual phenological monitoring essential for climate-adaptive management.

@@

\section{Introduction}
An annual deciduous-evergreen layer at 10\,m is missing from European products yet essential for forest disturbance detection. Copernicus DLT achieves 96.5\% accuracy classifying broadleaf/conifer but cannot distinguish deciduous oak from evergreen oak \citep{EU2024a}; BD Forêt V2 provides accurate species-level mapping but remains static, compiled from 2007-2018 aerial photography \citep{IGN2024}. Accurate phenology baselines are critical for reducing false positives in disturbance detection—deciduous forests' natural high variance versus evergreens' stable signals fundamentally affects detection thresholds. Species-specific vulnerabilities differ significantly between coniferous and deciduous forests, requiring precise phenological discrimination for early warning systems. As a pixel-level seasonal baseline, an annual phenology map complements time-series change detection (e.g., CCDC, BEAST, BFAST, LandTrendr) by constraining alerts during expected leaf-off/leaf-on periods and improving discrimination of genuine disturbances \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018}.

Foundation models in Earth observation have emerged as transformative technologies (2022-2025), with recent efforts such as SatMAE, Prithvi-EO, and AlphaEarth exemplifying this paradigm shift \citep{Cong2022,Szwarcman2024PrithviEO2,AlphaEarth2025}. These models, trained at scale on multi-temporal satellite imagery, demonstrate strong transfer to diverse downstream tasks. In particular, AlphaEarth Foundations releases annual 10\,m, 64-dimensional embeddings (2017–2024) in Earth Engine; vectors are unit-normalized and temporally consistent, enabling cosine-similarity change cues and low-shot transfer with linear probes or kNN \citep{AlphaEarth2025}.

Analysis-ready “Satellite Embedding V1” layers in Earth Engine allow practitioners to treat the 64-band embedding images as a ready-made feature stack—often replacing handcrafted indices or harmonic fits while reducing required labels \citep{Google2025SatelliteEmbeddingV1}. Early operational reports describe broad adoption across conservation and forestry programs \citep{DeepMind2025AlphaEarth}.

The versatility of embedding-based approaches extends across diverse Earth observation domains. In agricultural applications, self-supervised embeddings have shown competitive performance for crop and land-cover tasks \citep{Cong2022}, while general-purpose geospatial foundation models demonstrate strong transfer on multiple benchmarks and resolutions \citep{Cha2023Billion,Khanna2023DiffusionSat,Xie2024FoundationEffective}. SatMAE, in particular, adapts masked autoencoding with temporal/spectral encodings and independent masking, improving transfer by up to 14\% on downstream remote-sensing tasks \citep{Cong2022}. These systems handle challenging conditions (e.g., cloud cover and temporal gaps) and support efficient adaptation to new regions and tasks.

Disaster response and environmental monitoring increasingly leverage pre-trained representations for rapid deployment. Open geospatial foundation models and associated toolchains democratize access to high-performance geospatial analysis, enabling practitioners to build monitoring systems for food security, conservation, and disaster response with improved efficiency \citep{Szwarcman2024PrithviEO2,Xie2024FoundationEffective,NatureGFM2025}. Practical references and additional discussion are provided in Supplementary Section~S8.

Beyond architectures, dataset design measurably affects representation quality: phenology-informed, globally uniform pretraining datasets improve downstream ecological performance compared to calendar-season or urban-biased sampling \citep{Plekhanova2025SSL4Eco}. This supports our emphasis on phenological signals as primary discriminants in forest mapping.

Yet questions persist about performance versus domain-specific methods for specialized tasks like phenological classification, particularly regarding interpretability and cross-regional transferability where some studies show performance degradation in transfer scenarios \citep{Xie2024FoundationEffective}.

@@

Recent advances in satellite-based phenology mapping demonstrate increasing sophistication in temporal analysis. Time-series methods such as CCDC, BEAST, BFAST, and LandTrendr detect phenological changes and disturbances more reliably than simple seasonal baselines \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018}. Continental-scale phenology mapping from harmonized Landsat and Sentinel-2 and dedicated phenology indices further highlight the value of explicit temporal modeling \citep{Bolton2020,Li2023,Low2020}. However, foundation models like SatMAE, Prithvi, and AlphaEarth, while powerful, raise questions about temporal integration and cross-regional transferability for specialized tasks such as phenological classification \citep{Cong2022,Szwarcman2024PrithviEO2,AlphaEarth2025,Xie2024FoundationEffective}. Recent benchmarking efforts further show FMs can be markedly label-efficient yet emphasize the need for standardized, fair comparisons across tasks and regions \citep{Dionelis2024BenchmarkFM}. Our eco-region folds and weights follow this principle to ensure comparability between harmonics and embeddings.

Early preprints have begun using AlphaEarth embeddings directly for forest and disturbance tasks: Houriez et al. extend LANDFIRE vegetation types into Canada using simple classifiers on AEF features (\~81\% USA, 73\% Canada OA) \citep{Houriez2025AEFDataGen}; and a Siamese U-Net on AlphaEarth datasets reports strong burned-area mapping across regions \citep[][note: dataset branding vs 64D feature use varies]{Seydi2025AlphaEarthBurnedArea}.

Operational time-series mapping at country scale and phenology-driven evergreen mapping reinforce the role of temporal features in forest applications \citep{Inglada2017,Li2023,Low2020,Bolton2020}.

Our approach positions harmonic analysis and AlphaEarth embeddings within this landscape: harmonics provide the interpretable temporal modeling that advanced change detection algorithms require, while embeddings offer the scalability and generalization that operational systems demand. Evidence from national inventory studies likewise shows that deep pre-trained time-series features can outperform harmonic descriptors by large margins for species mapping \citep{Ishikawa2025NFI}. This comparison directly addresses the trade-offs between accuracy, interpretability, and operational feasibility that define modern Earth observation challenges.

\section{Data and Methods}

\subsection{Dataset composition and sampling}
We assemble a national supervised dataset of 14.1 million forest pixels by tiling mainland France into non-overlapping 2.5\,km\,$\times$\,2.5\,km units and sampling within these tiles across the eleven eco-regions (Figure~\ref{fig:training_tiles}). Labels indicate phenology class (deciduous vs evergreen) consolidated from national forest sources (e.g., inventory plots and mapped stands). Sampling is approximately balanced by eco-region using area-aware weights so that large Atlantic regions do not dominate smaller Mediterranean or Alpine regions. All models use the same sampled pixels, eco-region labels, and weights to ensure a fair comparison.

\subsection{Features}
\subsubsection{Harmonic features (on spectral indices)}
We derive four vegetation indices from Sentinel-2 time series at 10\,m: NDVI and EVI (canopy greenness/photosynthetic activity), NBR (moisture/char and structural change), and CRSWIR (vegetation water content and woody structure). After QA60/SCL cloud–snow masking and gap-filling on 10-day composites, each pixel's annual index signal \(x(t)\) is approximated by
\begin{equation}
  x(t) \approx C + \sum_{k=1}^{2} \big[ a_k \cos\!\big( \tfrac{2\pi k}{T} t \big) + b_k \sin\!\big( \tfrac{2\pi k}{T} t \big) \big],
\end{equation}
with period \(T=1\) year. We summarize the fit with interpretable descriptors: offset \(C\), first- and second-harmonic amplitudes \(A_k = \sqrt{a_k^2 + b_k^2}\), phases \(\varphi_k = \operatorname{atan2}(b_k, a_k)\) (reported via sine/cosine components for robustness), and residual variance. From these candidates we retain a compact set of 14 descriptors (“Harmonic-14”) that balances accuracy and simplicity (Supplement S1–S2). Using indices ties the features to physical vegetation processes—seasonal vigor, timing, and baseline moisture/structure—useful for distinguishing deciduous and evergreen behavior.

\subsubsection{AlphaEarth embeddings (what they represent and how they are obtained)}
Embeddings are compact numerical summaries of how a place looks and changes through the year—an annual “fingerprint” per 10\,m pixel. AlphaEarth provides, for each pixel and year, a 64-dimensional unit-length vector learned so that the vector can recreate multi-sensor observations over the chosen period. Training combines: (i) an encoder that ingests a local spatio‑temporal “video” of Earth observations (Sentinel‑2, Sentinel‑1, Landsat, and ancillary signals like topography and climate), (ii) implicit decoders that are conditioned on time and sensor parameters to reconstruct measurements, and (iii) objectives that stabilize and generalize the representation (consistency between teacher and student models, a uniformity constraint that spreads vectors on the unit hypersphere, and weak alignment to geocoded text). This process produces annual 64D embeddings that tend to align with physically meaningful signals: seasonal vigor and timing (akin to NDVI/EVI amplitudes and phases), moisture and burn sensitivity (akin to NBR), and baseline water/structure (akin to CRSWIR), while also carrying local spatial context and stable background information. Our similarity analysis supports this view: specific dimensions correlate with first-harmonic NDVI/NBR amplitude and with CRSWIR-related baselines.

\subsection{Training (including feature selection)}
We compare two compact, interpretable feature sets under identical data splits and regional weights: Harmonic‑14 and Emb‑14 (a 14‑dimension subset of the 64D vectors selected within training folds). We train a Random Forest classifier for its robustness and effectiveness on remote-sensing features, tuning hyperparameters on the training folds. To account for the larger share of deciduous pixels, we balance class frequencies during training.

\subsection{Evaluation protocol}
To avoid spatial information leakage, we evaluate with five-fold cross-validation where the 2.5\,km tiles are the grouping units: all pixels from a tile belong to exactly one fold. This prevents “seeing” neighboring pixels of a held‑out tile during training, which would otherwise inflate performance due to spatial autocorrelation. Folds are stratified by eco-region and we weight samples so each eco-region contributes proportionally to its forested area. We report overall accuracy (OA) and macro/weighted F1 at the fold level and also provide national aggregation with eco-region area weights.

\subsection{Embedding–harmonic similarity analysis (setup)}
We assess what embeddings encode relative to explicit phenology by linearly projecting each selected embedding dimension onto the full set of harmonic descriptors using standardized ridge regression. We evaluate out‑of‑sample, area‑weighted \(R^2\) by eco‑region and tile, and record the most correlated harmonic groups and normalized coefficients. Extended diagnostics appear in Supplement S3.

\subsection{Comparison with existing products}
We compare our annual phenology maps against Copernicus DLT (broadleaved vs coniferous) and BD Forêt V2 (deciduous vs evergreen). Comparisons are restricted to pixels mapped as forest and summarized at eco‑region and national scales; additional details and maps are provided in Supplement S9.
\textbf{Forest mask} — Final maps are constrained to the national forest extent using the IGN BD Forêt Beta mask \citep{IGN2024mask}, ensuring evaluation focuses on forested areas at 10\,m resolution.

\section{Results}
\subsection{National Comparison}
\begin{table}[H]
    \centering
    \begin{tabular}{lccc}
        \hline
        \textbf{Model} & \textbf{OA} & \textbf{F1\_macro} & \textbf{F1\_weighted} \\ \hline
        Emb-14 (baseline) & \textbf{0.924} (\(\pm\) 0.009) & \textbf{0.903} (\(\pm\) 0.006) & \textbf{0.926} (\(\pm\) 0.009) \\
        Harmonic-14 & 0.904 (\(\pm\) 0.007) & 0.874 (\(\pm\) 0.011) & 0.905 (\(\pm\) 0.006) \\
        \hline
    \end{tabular}
    \caption{National performance under identical eco-region folds and weights (mean ± sd across folds).}
    \label{tab:national_comparison}
\end{table}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Dataset tiles (training and validation) distributed across France's eleven eco-regions, ensuring representation of diverse phenological patterns from oceanic to mediterranean climates. Each 2.5\,km tile contains multiple labeled pixels for cross-validation.}
    \label{fig:training_tiles}
\end{figure}

@@

Performance varied meaningfully across eco-regions, reflecting forest composition and phenological complexity (Table \ref{tab:eco_region_comparison}). Table \ref{tab:eco_region_comparison} presents results for both the full 64-embedding configuration (Emb-64) and the 14 selected embeddings (Emb-14), alongside harmonics. We selected 14 harmonic descriptors using a performance-guided procedure within cross-validation, and used the same number for embeddings to ensure a fair comparison. While Emb-14 demonstrates strong performance, Emb-64 achieves even higher accuracies across all regions, demonstrating the power of the complete embedding representation (see Supplementary Materials for selection details). Embeddings achieved superior accuracy in every region, with largest improvements in challenging environments: Mediterranean (+2.5 percentage points) and Corsica (+4.0 percentage points). These gains highlight embeddings' ability to capture subtle patterns that harmonic analysis struggles to encode.
\begin{table}[H]
\centering
\begin{tabular}{lcccc}
\hline
\textbf{Eco-region} & \textbf{OA (Harmonic-14)} & \textbf{OA (Emb-64)} & \textbf{OA (Emb-14)} & \textbf{Samples} \\ \hline
Alps & 0.869 & \textbf{0.906} & 0.888 & 0.72\,M \\
Central Massif & 0.895 & \textbf{0.935} & 0.927 & 1.93\,M \\
Corsica & 0.654 & \textbf{0.694} & 0.662 & 0.35\,M \\
Greater Crystalline and Oceanic West & 0.873 & \textbf{0.946} & 0.931 & 0.49\,M \\
Greater Semi-Continental East & 0.952 & \textbf{0.967} & 0.960 & 2.87\,M \\
Jura & 0.882 & \textbf{0.936} & 0.920 & 0.20\,M \\
Mediterranean & 0.811 & \textbf{0.836} & 0.803 & 1.46\,M \\
Oceanic Southwest & 0.925 & \textbf{0.959} & 0.953 & 2.66\,M \\
Pyrenees & 0.930 & \textbf{0.965} & 0.947 & 0.56\,M \\
Semi-Oceanic North Center & 0.937 & \textbf{0.966} & 0.958 & 2.46\,M \\
Vosges & 0.880 & \textbf{0.908} & 0.900 & 0.40\,M \\ \hline
\end{tabular}
\caption{Eco-region accuracy comparison. Bold values indicate best performance per region. Largest embedding gains occur in Mediterranean environments where phenological boundaries blur.}
\label{tab:eco_region_comparison}
\end{table}

@@

% (Removed duplicate results subsection on product comparison; consolidated under the later section.)

@@

Figure \ref{fig:national_map} presents our 2023 deciduous-evergreen classification of French forests using the embedding-based model (Emb-14), which achieved superior national accuracy. The map reveals patterns invisible in traditional forest classifications: sharp deciduous-evergreen boundaries along elevation gradients, maritime pine dominance along the Atlantic coast, and complex mosaics in managed forests.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/France_Phenology_Map_RF.png}
    \caption{France's 2023 10\,m deciduous-evergreen forest map produced using the embedding-based model (Emb-14) reveals phenological patterns invisible in leaf-type classifications. Orange areas indicate deciduous forests, cyan shows evergreen forests. The embedding approach provides improved boundary delineation in mixed stands compared to traditional methods.}
    \label{fig:national_map}
\end{figure}

@@

\section{Embedding–harmonic similarity analysis}

See Data and Methods for the setup (ridge projection onto the harmonic base; grouped cross-validation by tile; weighted out-of-sample $R^2$); extended diagnostics and outputs are reported in Supplement S3.

@@

Table \ref{tab:similarity_ecoregion} summarizes the updated analysis. Overall linear alignment remains poor (mean out-of-sample R² < 0 in every region), yet individual embedding dimensions show moderate correlations with specific phenological descriptors (notably first-harmonic amplitudes and CRSWIR offsets), revealing interpretable structure.

\begin{table}[H]
\centering
\begin{tabular}{lcc}
\hline
\textbf{Eco-region} & \textbf{Mean R²} & \textbf{Top group (|r| mean)} \\ \hline
Vosges & -0.06 & CRSWIR: offset (0.68) \\
Alps & -0.31 & NDVI: amplitude\_h1 (0.54) \\
Greater Crystalline and Oceanic West & -0.35 & NBR: amplitude\_h1 (0.66) \\
Semi-Oceanic North Center & -0.33 & NBR: amplitude\_h1 (0.54) \\
Central Massif & -0.32 & CRSWIR: offset (0.65) \\
Mediterranean & -0.49 & NBR: amplitude\_h1 (0.62) \\
Oceanic Southwest & -0.87 & NBR: amplitude\_h1 (0.61) \\
Greater Semi-Continental East & -0.56 & NBR: amplitude\_h1 (0.52) \\
Corsica & -0.72 & NDVI: amplitude\_h1 (0.55) \\
Pyrenees & -0.90 & NDVI: amplitude\_h1 (0.54) \\
Jura & -0.72 & CRSWIR: offset (0.54) \\ \hline
\end{tabular}
\caption{Embedding–harmonic linear alignment by eco-region (out-of-sample mean R², top correlated harmonic group with mean absolute correlation). Full top-3 per region in Supplement~S3.}
\label{tab:similarity_ecoregion}
\end{table}

@@

Despite the poor global linear fit, three robust, interpretable patterns emerge across regions:

\textbf{Seasonal amplitude (first harmonic)}: Multiple embedding dimensions align with NDVI/NBR amplitude\_h1 (|r| ≈ 0.5–0.7 in several regions), capturing overall seasonal strength—the primary discriminator between deciduous and evergreen forests.

\textbf{Baseline spectral properties}: CRSWIR offset shows consistent, moderate alignment (|r| often 0.5–0.7), indicating embeddings encode baseline water content/structure beyond purely temporal variation.

\textbf{Regional specialization}: Mountain regions (Vosges, Jura, Central Massif) emphasize CRSWIR offsets; Atlantic/Western regions emphasize NBR amplitude/offset; Mediterranean adds NDVI phase\_h1, consistent with complex phenological timing under drought stress.

Component-wise, first-harmonic amplitudes dominate (mean |r| ≈ 0.50 across 16 region-usages), followed by offsets (≈0.44 across 19). This pattern aligns with phenology (seasonal vigor) and baseline water/structure being the primary axes of linear alignment.

These findings support a physically grounded view of the embeddings as annual fingerprints: they encode seasonal vigor, baseline moisture/structure, and some timing information, alongside additional spatial context not captured by harmonic decomposition alone. Region-level summaries and coefficient diagnostics are provided in Supplement S3.
